Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-10-31 forecast/paper/supplement-text.tex
Another intermediate commit on supp

\section{Aggregation With Geometric Mean}

In this section, we consider using the geometric mean instead of the usual
(arithmetic) mean when aggregating the weighted interval score (WIS) across
location-time pairs.  There are two reasons why using the geometric mean may be
desirable.

\begin{enumerate}
\item WIS is right-skewed, being bounded below by zero and having occasional 
  very large values.  Figure \ref{fig:wis-densities} illustrates that the
  densities appear roughly log-Gaussian.  The geometric mean is a natural choice   
  in such a context since it can be viewed as a measure of centrality on the log
  scale (it is the exponential of the arithmetic mean of log values).
\item In the main paper, we report the ratio of the mean WIS of a forecaster to
  the mean WIS of the baseline forecaster. Another choice could be to take the
  mean of the ratio of WIS values for the two methods. This latter choice would
  penalize a method less for doing poorly where the baseline forecaster also
  does poorly.\footnote{In a sense, this is implicitly estimating a
    nonparametric space-time effect for forecaster error, and assuming that has
    a shared, multiplicative contribution to forecaster errors.  That is, if one
    imagines that a forecaster's WIS is composed of multiplicative space-time
    effects $S_{\ell,t}$ shared across all forecasters,
    $\WIS(F_{\ell,t,f},Y_{\ell,t}) = S_{\ell,t} E_{f,t}$ with $E_{f,t}$ a 
    forecaster-specific error, then taking the ratio of individual WIS values
    cancels these space-time effects.}  Using instead the geometric
  mean makes the order of aggregation and scaling immaterial since the ratio of
  geometric means is the same as the geometric mean of ratios.
\end{enumerate}

@@

\section{Comparing With the COVID-19 Forecast Hub}

@@

Our goal in this section is to compare the AR model and indicator models to
those in the Hub, in terms of forecast errors aggregated over the same forecast
tasks, to give a clear sense for how robust and effective the models we choose
to investigate in the paper are relative to those in common operational use.
This was prompted by a question from an anonymous reviewer of this paper, who
asked why we chose to build our analysis of indicator utility around the AR
model in the first place, and why we did not investigate others (say, the SIR
model or more complex mechanistic models of disease transmission) that have
occupied more of the spotlight over the course of the pandemic.  The analysis
presented here corroborates the claim that an AR model is simple, and yet,
properly trained---using a quantile loss to directly estimate multiple
conditional quantiles, a trailing training window of 21 days, pooling across all
locations jointly, and fitting to case rates rather than counts---all as we do
in the models in the main paper, it can be robust and effective, performing 
competitively with the top models from COVID-19 Forecast Hub.

The closest forecast target in the Hub to that used in this paper are
state-level case incidence over an epiweek---defined by the sum of new case
counts reported between a Sunday and the following Saturday (inclusive). Our
target is 7-day trailing average of COVID-19 case incidence rates at the HRR
level, which is different in three regards:

\begin{enumerate}
\item temporal resolution (daily versus weekly);
\item geographic resolution (HRRs versus states); and 
\item scale (rates versus counts).
\end{enumerate}

While the first and third of these differences can be easily addressed post 
hoc---meaning, we can take always take our model's outputs and multiply them by
7 in order to track the incidence over any given trailing week, and rescale them
per-location by population to put them on the count scale---the second
difference is not easy to adjust post hoc, because a quantile is not linear
operator (a quantile of a linear combination of random variables is not simply
the linear combination of their quantiles, but rather, depends intricately on
the correlations between random variables).

Therefore, to make a comparison to models in the Hub in as direct a way as
possible, we retrained our models over the same forecast period as in the main
paper, and with the same general setup entirely, except at the state rather than 
HRR level.  We then rescaled them post hoc to account for the different temporal
resolution and the rate versus count scale (first and third items on the above
list).  The results are given in Figure \ref{fig:compare-to-hub}.  The
evaluation was carried out exactly as in the main paper, and the figure displays
both mean WIS and geometric mean WIS, as a function of ahead.  Furthermore, to
account for missingness (not all teams submitted forecasts to the Hub for all
locations and ahead values for the entire period), we first dropped any
forecaster than submitted for less than 6 weeks, and then restricted the
aggregation metrics (mean or geometric mean) to errors from commonly available
forecast tasks.

By either metric, mean and geometric mean WIS, relative to baseline, we can see
in Figure \ref{fig:compare-to-hub} that the AR model examined in this paper is 
competitive with top models in the Hub, even outperforming the Forecast Hub
ensemble model for smaller ahead values.  The same general conclusion can be
drawn for the indicators models as well.  However, interestingly, a close
inspection reveals that the AR model here is for the most part in the ``middle
of the pack'' when compared to the indicator models, and only the Google-AA
model offers clear improvement over AR for all aheads.  This is likely due to
the fact that at the state level, the signal-to-noise ratio (SNR) is generally
higher, and AR model provides a higher standard (on which to expect improvement
using an auxiliary indicator).  At the HRR level, with lower SNR, using
indicators as simple additional linear features in the AR model provides an
important enough variance reduction

% Essentially, for each state, teams are asked to predict the total number of
% cases that will be recorded over the week from Monday to Sunday. Forecasts must
% be submitted by Tuesday for the current Sunday as well as the next three
% Sundays. This corresponds to producing forecasts for 5, 12, 19, and 26 days
% ahead. Some teams choose to submit their forecasts a few days earlier, but
% always for the same targets.

To properly compare the models presented in this paper with those submitted to
the Hub, we limit ourselves to the same set of forecast dates and states, and
we predict case incidence over a week by: (1) predicting the 7-day trailing
average case rate (as in the manuscript);
(2) multiplying by 7 to give the total over a week; (3) multiplying by the state
population as reported in the 2019 US Census Population Estimate (available in
the Covidcast R package \cite{CovidcastR}). Finally, we perform the same
evaluations as in the manuscript Figure 3 and also using geometric mean as
described above.
For the purpose of the comparison, we show only the AR model along with
submissions collected by the Forecast Hub in Figure \ref{fig:compare-to-hub}.
Since not all teams submitted forecasts for the entire period, we
only display those which submitted for at least 6 weeks.

We also are undertook the same analysis shown in Figure 3 in the manuscript and Figure
\ref{fig:fcast-adjusted} above for all models
(the AR, as well as the indicator-assisted models). This comparison is shown in Figure
\ref{fig:state-level-performance}. 
%
TODO: some text about why indicators don't help as much.

@@

In the Introduction of the main paper, we give some reasons that we avoid making
formal statements about statistical significance, preferring instead to examine
the stability of our results in different contexts. There are strong reasons to
avoid model-based significance tests because the necessary assumptions about
stationarity, independence, and the  model being true (or at least approximately
true) are certainly violated. With those caveats in mind, we undertake two
relatively assumption-lean investigations in this section. The first is a
sign-test for whether the difference between the AR model’s relative WIS  and
each other model’s relative WIS is centered at zero. By ``relative WIS'' we mean
scaled by the strawman as displayed in Figure 3. To mitigate the dependence
across time (which intuitively seems to matter more than that across space), we
computed these tests in a stratified way, where for each forecast date we run a
sign test on the scaled errors between two models over all 440 counties. The
results are plotted as histograms in Figure \ref{fig:sign-test}. In this case,
we use the total relative WIS over all aheads, but the histograms are largely
similar for individual target horizons. If there were no difference, we would
expect to see a uniform distribution. However, for all indicator-assisted
models, we see many more small p-values than would be expected if the null
hypothesis (that the AR model is better) were true.

Another relatively assumption-lean method of testing for differences in forecast
accuracy is the Diebold-Mariano test \cite{Diebold:2002, Diebold:2015,
Harvey:1997}. Essentially, the differences between forecast losses are assumed
to have a constant mean and a covariance that depends on time. Under these
conditions, the asymptotic distribution for the standardized mean of the
differences is limiting normal provided that a heteroskedasticity and
autocorrelation robust estimate of the variance is used. Using the loss as
weighted interval score across all HRRs and horizons (7 to 21 days ahead), we
perform the DM test using both the mean relative to the strawman (as reported in
the manuscript) and the geometric mean relative to the strawman as described
above. The first two rows of Table \ref{tab:dm-test} displays p-values for the
test that the indicator-assisted model is no-better than the AR model. Only the
CHNG-CLI model's p-value exceeds conventional statistical significance
thresholds.

@@

As explained in Section 2.B. of the main paper, a (somewhat cynical) hypothesis
for why we see benefits in forecasting and hotspot prediction is that the
indicators are not actually providing useful information but they are instead
acting as a sort of ``implicit regularization,'' leading to shrinkage on the
autoregressive coefficients and therefore to less volatile predictions.  To
investigate this hypothesis, we consider fitting ``noise features'' that in
truth should have zero coefficients.  Recall (from the main paper) that at each
forecast date, we train a model on 6,426 location-time pairs.  Indicator models
are based on six features, corresponding to the three autoregressive terms and
the three lagged indicator values.  To form noise indicator features, we replace
their values with those from a randomly chosen time-space pair (while keeping
the autoregressive features fixed).  In particular, at each location $\ell$ and
time $t$, for the forecasting task we replace the triplet $(X_{\ell,t},
X_{\ell,t-7}, X_{\ell,t-14})$ in Eq. (3) of the main paper with the triplet
$(X_{\ell^*,t^*}, X_{\ell^*,t^*-7}, X_{\ell^*,t^*-14})$, where $(\ell^*,t^*)$ is
a location-time pair sampled with replacement from the 6,426 location-time
pairs.  Likewise in the hotspot prediction task, we replace the triplet
$(X_{\ell,t}^\Delta, X_{\ell,t-7}^\Delta, X_{\ell,t-14}^\Delta)$ in Eq. (5) of
the main paper with $(X_{\ell^*,t^*}^\Delta, X_{\ell^*,t^*-7}^\Delta,
X_{\ell^*,t^*-14}^\Delta)$.  Figures
\ref{fig:fcast-booted}--\ref{fig:hot-booted} show the results.  No method
exhibits a noticeable performance gain over the \ar~method, leading us to
dismiss the implicit regularization hypothesis.

@@

In this section we provide extra details about the upswing / flat / downswing
analysis described in the main text. Figure \ref{fig:upswing-summary} shows the
overall results, examining the average difference $\WIS(AR) - \WIS(F)$ in
period. Figure \ref{fig:hotspots-upswing-downswing} shows the same information
for the hotspot task. On average, during downswings and flat periods, the
indicator-assisted models have lower classification error and higher log
likelihood than the AR model. For hotspots, both Google-AA and CTIS-CLIIC
perform better than the AR model during upswings, in contrast to the forecasting
task, where only Google-AA improves. For a related analysis, Figure
\ref{fig:cor-wis-ratio} shows histograms of the Spearman correlation (Spearman's
$\rho$, a rank-based measure of association) between the $\WIS(F)/\WIS(AR)$ and
the magnitude of the swing. Again we see that case rate increases are positively
related to diminished performance of the indicator models.

@@

Figure \ref{fig:errs-in-space} examines the spatial behavior over the entire
period of the indicator-assisted models relative to the AR. For ease of
comparison, we show the percent improvement in each HRR. Negative numbers (blue)
mean that the indicator helped while positives (red) mean that the indicator
hurt forecasting performance. The clear pattern is that in most HRRs, the
indicators improved performance, though usually by small amounts (2.5\%--10\%).
In some isolated HRRs, performance was markedly worse, though there does not
appear to be any particular pattern to these locations.

>>>>

\section{Robust Aggregation}

In this section, we consider using the geometric mean instead of the usual
(arithmetic) mean when aggregating the weighted interval score (WIS) across
location-time pairs.  Aside from the geometric mean being generally more robust
to large values, there are two reasons why using it may be desirable.

\begin{enumerate}
\item WIS is right-skewed, being bounded below by zero and having occasional 
  very large values.  Figure \ref{fig:wis-densities} illustrates that the
  densities appear roughly log-Gaussian.  The geometric mean is a natural choice   
  in such a context since it can be viewed as a measure of centrality on the log
  scale (it is the exponential of the arithmetic mean of log values).
\item In the main paper, we report the ratio of the mean WIS of a forecaster to
  the mean WIS of the baseline forecaster. Another choice could be to take the
  mean of the ratio of WIS values for the two methods. This latter choice would
  penalize a method less for doing poorly where the baseline forecaster also
  does poorly.\footnote{In a sense, this is implicitly estimating a
    nonparametric space-time effect for forecaster error, and assuming that has
    a shared, multiplicative contribution to forecaster errors.  That is, if one
    imagines that a forecaster's WIS is composed of multiplicative space-time
    effects $S_{\ell,t}$ shared across all forecasters,
    \smash{$\WIS(F_{\ell,t,f},Y_{\ell,t}) = S_{\ell,t} E_{f,t}$} with
    \smash{$E_{f,t}$} a forecaster-specific error, then taking the ratio of
    individual WIS values cancels these space-time effects.}  Using instead the
  geometric mean makes the order of aggregation and scaling immaterial since the
  ratio of geometric means is the same as the geometric mean of ratios.
\end{enumerate}

@@

\section{Comparing COVID-19 Forecast Hub Models}

@@

Our goal in this section is to compare the AR model and indicator models to
those in the Hub, in terms of forecast errors aggregated over the same forecast
tasks, to give a clear sense for how robust and effective the models we choose
to investigate in the paper are relative to those in common operational use.
This was prompted by a question from an anonymous reviewer of this paper, who
asked why we chose to build our analysis of indicator utility around the AR
model in the first place, and why we did not build it around others (say, the
SIR model or more complex mechanistic models of disease transmission) that have 
occupied more of the spotlight over the course of the pandemic.  The analysis
presented here corroborates the claim that, while simple, the AR model, properly
trained---using a quantile loss to directly estimate multiple conditional
quantiles, a trailing training window of 21 days, pooling across all locations
jointly, and fitting to case rates rather than counts (as we do in all our
models in the main paper)---can be robust and effective, performing
competitively many of the top models from COVID-19 Forecast Hub, including the
Hub ensemble model.

The closest forecast target in the Hub to that used in the main paper is 
state-level case incidence over an epiweek---defined by the sum of new case
counts reported between a Sunday and the following Saturday (inclusive). Our
forecast target, recall, is a 7-day trailing average of COVID-19 case incidence
rates at the HRR level, which is different in three regards:

\begin{enumerate}
\item temporal resolution (daily versus weekly);
\item geographic resolution (HRRs versus states); 
\item scale (rates versus counts).
\end{enumerate}

While the first and third of these differences could be easily addressed post 
hoc---meaning, we can take always take our model's output and multiply it by 
7 in order to track the incidence over any given trailing week, and rescale it
per location by population to bring it to the count scale---the second
difference is not easy to adjust post hoc due to nonlinearity of the quantiles
(a quantile of a linear combination of random variables is not simply the linear
combination of their quantiles, but rather, depends intricately on the
correlations between the random variables).

Therefore, to make a comparison to models in the Hub as direct as possible, we
retrained our models over the same forecast period as in the main paper, and
with the same general setup entirely, except at the state rather than HRR level.
We then rescaled them post hoc to account for the different temporal resolution
and the rate versus count scale (first and third points in the above list).  The
results are given in Figure \ref{fig:compare-to-hub}.  The evaluation was
carried out exactly as in the main paper, and the figure displays both mean WIS
and geometric mean WIS, as a function of ahead, relative to the baseline model.
Furthermore, to account for missingness (not all teams submitted forecasts to
the Hub for all locations and ahead values for the entire period), we first
dropped any forecaster than submitted for less than 6 weeks, and then restricted
the aggregation metrics (mean or geometric mean) to errors from commonly
available forecast tasks.

By either metric, mean or geometric mean WIS relative to baseline, we can see  
in Figure \ref{fig:compare-to-hub} that the AR model examined in this paper is 
competitive with top models in the Hub, even outperforming the Hub ensemble
model for smaller ahead values.  The same general conclusion can be drawn for
the indicators models as well.  However, interestingly, a close inspection
reveals that the AR model here is for the most part in the ``middle of the
pack'' when compared to the indicator models, and only the Google-AA model
offers clear improvement over AR for all aheads.  This is likely due to the fact
that at the state level, the signal-to-noise ratio (SNR) is generally higher,
and AR model provides a higher standard (on which to expect improvement using an 
auxiliary indicator), since it is able to extract a clearer signal from past
lags of case rates.  At the HRR level, with lower SNR, using indicators as
simple additional linear features in the AR model probably leads to a variance
reduction that is enough to boost accuracy, but at the state level, perhaps more
sophisticated modeling techniques are needed to extract comparable value from
some of the indicators.

@@

In the introduction of the main manuscript, we gave some reasons that we avoid
making formal statements about statistical significance, preferring instead to
examine the stability of our results in different contexts.  There are strong
reasons to avoid model-based significance tests because the necessary
assumptions about stationarity, independence, and the model being true (or at
least approximately true) are certainly violated.  With those caveats in mind,
we undertake two relatively assumption-lean investigations in this section.  The
first is a sign test for whether the difference between the AR model’s relative
WIS and each other model’s relative WIS is centered at zero.  (Relative WIS here
means scaled by the WIS of the baseline model.)  To mitigate the dependence 
across time (which intuitively seems to matter more than that across space), we
computed these tests in a stratified way, where for each forecast date we run a
sign test on the scaled errors between two models over all 306 HRRs.  The
results are plotted as histograms in Figure \ref{fig:sign-test}.  For this
figure, we use the total relative WIS over all aheads, but the histograms are
largely similar for individual target horizons. If there were no difference, we
would expect to see a uniform distribution. However, for each indicator model, 
we see many more small p-values than would be expected if the null hypothesis
(that the AR model is better) were true.

Another relatively assumption-lean method of testing for differences in forecast
accuracy is the Diebold-Mariano (DM) test \cite{Diebold:2002, Diebold:2015,
Harvey:1997}. Essentially, the differences between forecast errors are assumed
to have a constant mean and a covariance that depends on time.  Under these
conditions, the asymptotic distribution for the standardized mean of the
differences is limiting normal provided that a heteroskedasticity and
autocorrelation robust estimate of the variance is used.  Using the error as WIS
across all HRRs and horizons (7 to 21 days ahead), we perform the DM test using
both the mean relative to the baseline (as reported in the manuscript) and the
geometric mean relative to the baseline as described above.  The first two rows
of Table \ref{tab:dm-test} displays p-values for the test that each indicator
model is no better than the AR model.  In only a few instances---geometric mean
and mean for the CHNG-CLI model, and mean for the CHNG-COVID model---do the
p-values exceed conventional statistical significance thresholds.

@@

As explained in Section 2.B of the main paper, a (somewhat cynical) hypothesis
for why we see benefits in forecasting and hotspot prediction is that the
indicators are not actually providing useful information but they are instead
acting as a sort of ``implicit regularization,'' leading to shrinkage on the
autoregressive coefficients and therefore to less volatile predictions.  To
investigate this hypothesis, we consider fitting ``noise features'' that in
truth have zero relationship to the response.  Recall (from the main paper) that
at each forecast date, we train a model on 6,426 location-time pairs.  Indicator
models are based on six features, corresponding to the three autoregressive
terms and the three lagged indicator values.  To form noise indicator features,
we replace their values with those from a randomly chosen time-space pair (while
keeping the autoregressive features fixed).  In particular, at each location
$\ell$ and time $t$, for the forecasting task we replace the triplet
\smash{$(X_{\ell,t}, X_{\ell,t-7}, X_{\ell,t-14})$} in Eq.\ (3) of the main
paper with the triplet \smash{$(X_{\ell^*,t^*}, X_{\ell^*,t^*-7},
  X_{\ell^*,t^*-14})$}, where $(\ell^*,t^*)$ is a location-time pair sampled with
replacement from the 6,426 total location-time pairs.  Likewise in the hotspot 
prediction task, we replace the triplet \smash{$(X_{\ell,t}^\Delta,
  X_{\ell,t-7}^\Delta, X_{\ell,t-14}^\Delta)$} in Eq.\ (5) of the main paper with
\smash{$(X_{\ell^*,t^*}^\Delta, X_{\ell^*,t^*-7}^\Delta,
  X_{\ell^*,t^*-14}^\Delta)$}.  Figures
\ref{fig:fcast-booted}--\ref{fig:hot-booted} show the results.  No method
exhibits a noticeable performance gain over the \ar~method, leading us to
dismiss the implicit regularization hypothesis.

@@

In this section, we provide extra details about the up/down/flat analysis
described in the main text. Figure \ref{fig:upswing-summary} shows the overall
results, examining the average difference $\WIS(AR) - \WIS(F)$ in period. Figure
\ref{fig:hotspots-upswing-downswing} shows the same information for the hotspot
task. On average, during downswings and flat periods, the indicator models have
lower classification error and higher log likelihood than the AR model. For
hotspots, both Google-AA and CTIS-CLIIC perform better than the AR model during
upswings, in contrast to the forecasting task, where only Google-AA
improves. For a related analysis, Figure \ref{fig:cor-wis-ratio} shows
histograms of the Spearman correlation (Spearman's $\rho$, a rank-based measure
of association) between the $\WIS(F)/\WIS(AR)$ and the magnitude of the
swing. Again we see that case rate increases are positively related to
diminished performance of the indicator models.

@@

Figure \ref{fig:errs-in-space} examines the spatial behavior over the entire
period of the indicator models relative to the AR. For ease of
comparison, we show the percent improvement in each HRR. Negative numbers (blue)
mean that the indicator helped while positives (red) mean that the indicator
hurt forecasting performance. The clear pattern is that in most HRRs, the
indicators improved performance, though usually by small amounts (2.5\%--10\%).
In some isolated HRRs, performance was markedly worse, though there does not
appear to be any particular pattern to these locations.
